Gaussian Process regression for astronomical time-series

### 2022-09-14 src/tex/ms.tex
[showyourwork] overleaf sync

Now, instead of assuming that the covariance matrix takes this very specific form, let us allow a more flexible covariance model:
\begin{equation}
  K_{ij} = k(t_i,t_j,\hyperparams) + \delta_{ij} \sigma_i^2,
\end{equation}
where $k$ is a \emph{covariance function} or \emph{kernel}, controlled by parameters $\hyperparams$. The result is a GP, and its likelihood is still given by Equation~(\ref{eq:gp_like}). Depending on the choice of kernel and parameters, the covariance matrix can now have non-zero off-diagonal elements, allowing us to explicitly model correlated noise or stochastic signals in the data.
Note that the likelihood depends not only on the argument of the exponential, but also on the determinant of the covariance matrix, which therefore needs to be evaluated explicitly. This term acts as a built-in Occam's razor, automatically penalising more complex models.

@@

% Assigned: DFM

% \paragraph{Kernels} Selecting a kernel function is how the user builds physical domain knowledge into the Gaussian process model. In this section, we will both try to dispel some common misconceptions about how to choose covariance functions and their parameters, and provide some practical guidelines for how to make these modelling choices in practice, highlighting specific examples from the astronomical literature. Estimated length: 5-10 pages

% Question: will this also be the section where the most commonly used kernel functions in astrophysics are introduced or will that be part of section1? This will affect the relative length of the two sections, but overall I'd expect it to be 15-20 pages for both sections

% \paragraph{Tuning} This section will cover the process of "training" a Gaussian process model, i.e. tuning the parameters of the covariance function and, if included, the mean function. Compared to standard texts from other fields, this topic will be covered in more depth because parameter inference (as opposed to interpolation or prediction) is one of the most common applications of Gaussian processes in astronomy. Once again we will aim to dispel some misconceptions and provide practical advice. This will also be where we illustrate how a Gaussian process model can be embedded in a Bayesian inference framework using posterior sampling, and point out the importance of marginalising over nuisance parameters. Estimated length: 5-10 pages

@@

The kernel function can be any positive semi-definite scalar function, but some will be more useful than others.
% A commonly (over-)used kernel function is the ``exponential-squared'' or ``radial basis function'' kernel
% \begin{equation}
%   k(t_i,\,t_j;\,\hyperparams) = \sigma^2\,\exp\left[-\frac{(t_i - t_j)^2}{2\,\tau^2}\right]
% \end{equation}
% where the ``hyperparameters'' $\hyperparams$ are $\sigma$ and $\tau$.
% In this case, the hyperparameters specify the typical amplitude $\sigma$ and timescale $\tau$ of the variance of the process.
% SOMEFIGURE shows how changing these parameters affect our GP model.
% There exist many other commonly used kernel functions, some of which are listed in SOMETABLE.
% This table is far from comprehensive, and importantly, the sum or product of two valid kernel functions is itself also a valid kernel, so it is possible (and common) to build extremely flexible and expressive kernel functions specific to your use case.
Given this large decision space, you may be wondering how to choose the right kernel for your specific problem.
If you're very lucky, you may be able to motivate your model using physics.
This can be approached from two directions.
On one hand, some commonly used kernel function have a specific physical interpretation---for example, the solution to a stochastic ordinary differential equation---that can be used to motivate their use.
On the other hand, if your physical model is specified by its \emph{power spectrum}, this can recast as a GP model with a specific kernel function. \suz{Say something about the number of times a function is differentiable? That can be a useful consideration. But perhaps already mentioned later.}

@@

\dan{Say something here about assessing Gaussianity.}

@@

Other common use cases, like the time delay example in \autoref{sec:quasar}, primarily require good predictive performance for the GP model.
In these cases, methods like cross-validation of posterior predictive assessment \dan{add link to Gelman paper here} can be used to evaluate different choices of kernel function.

@@

First, using the full dataset, we compute the Bayesian evidence integral for each of these model choices using a nested sampling algorithm implemented in the \textsf{jaxns} package \dan{add citation}.
The Bayesian evidence integral is defined as:
\begin{equation}
  Z(H) \equiv p(\ydata\,|\,H) = \int p(\meanparams,\hyperparams\,|\,H)\,p(\ydata\,|\,\meanparams,\hyperparams,H)\,\mathrm{d}\meanparams\,\mathrm{d}\hyperparams
\end{equation}
where $H$ indicates the modeling choices, in this case the choice of kernel function.
For better or worse\footnote{A discussion of the relative merits of different model selection algorithms is beyond the scope of this review. \suz{Sure but can we link to a more in-depth discussion of the topic for interested readers?}}, the Bayesian evidence is frequently used in the astrophysics literature as an ingredient in model selection procedures.
Nested sampling is an algorithm for numerically estimating this integral, and we apply it to produce estimates of $Z$ for each choice of kernel $H$, and plot these results in the middle panel of \autoref{fig:assessment}.
In this figure, it is clear that the squared exponential and rational quadratic kernels are indistinguishable under this metric, while the Mat\'ern-3/2 kernel is somewhat disfavoured.

@@

So far, we have primarily discussed the benefits of GP models for astrophysics, but in this section we aim to highlight some challenges faced when using GPs in practice.
The most significant challenge that has so far limited broader adoption GPs for astrophysics is their computational cost and scaling.

@@

The most widely-used algorithms for scalable exact GPR in time domain astrophysics are based---often not explicitly---on stochastic differential equations \citep[SDEs;][]{Sarkka:2019}, and typically require that the data be ``sortable''.
As an early example in the astrophysics literature, \citet{pr95} presented an algorithm for GPR with linear scaling with the size of the dataset, when using an exponential kernel with unevenly sampled time series data.
This method was later generalized to a wider range of covariance functions \citep{ambikasaran2015generalized, celerite, 2020AJ....160..240G, 2020A&A...638A..95D, 2022A&A...659A.182D}, and the \project{celerite}\footnote{\url{https://github.com/dfm/celerite}} and \project{S+LEAF}\footnote{\url{https://obswww.unige.ch/~delisle/spleaf/doc}} implementations have been widely-used for time domain astronomy. \dan{Add some citations to show the range.}
% \suz{yes. I take it you're reluctant to cite the S+LEAF papers?}.
A qualitatively different, and independently developed approach to implementing these models uses a state-space representation for the underlying SDE to enable scalable inference \citep{2014ApJ...788...33K,  2021RNAAS...5..107J, 2022arXiv220709327M} \dan{Maybe add some citations of the use of these methods}.

@@

Normalization.

\subsection{Non-Gaussianity}

>>>>

Now, instead of assuming that the covariance matrix takes this very specific form, let us allow a more flexible covariance model:
\begin{equation}
  K_{ij} = k(t_i,t_j,\hyperparams) + \delta_{ij} \sigma_i^2,
\end{equation}
where $k$ is a \emph{covariance function} or \emph{kernel}, controlled by parameters $\hyperparams$. The result is a GP, and its likelihood is still given by Equation~(\ref{eq:gp_like}). Depending on the choice of kernel and parameters, the covariance matrix can now have non-zero off-diagonal elements, allowing us to explicitly model correlated noise or stochastic signals in the data.
Note that the likelihood depends not only on the argument of the exponential, but also on the determinant of the covariance matrix, which therefore needs to be evaluated explicitly. This term acts as a built-in Occam's razor, automatically penalising more complex models.
\dan{We should probably define the determinant notation here? I don't think we really clearly define our matrix notation anywhere else.}

@@

@@

The kernel function can be any positive semi-definite scalar function, but some will be more useful than others.
Given this large decision space, you may be wondering how to choose the right kernel for your specific problem.
If you're very lucky, you may be able to motivate your model using physics.
This can be approached from two directions.
On one hand, some commonly used kernel function have a specific physical interpretation---for example, the solution to a stochastic ordinary differential equation---that can be used to motivate their use.
On the other hand, if your physical model is specified by its \emph{power spectrum}, this can recast as a GP model with a specific kernel function. \suz{Say something about the number of times a function is differentiable? That can be a useful consideration. But perhaps already mentioned later.}

@@

\dan{Say something here about assessing Gaussianity, and more generally model checking (rather than comparison; posterior predictive checks).}

@@

Other common use cases, like the time delay example in \autoref{sec:quasar}, primarily require good predictive performance for the GP model.
In these cases, methods like cross-validation \citep[e.g.][]{gelfand1992model} or posterior predictive assessment \citep[e.g.][]{gelman1996posterior} can be used to evaluate different choices of kernel function.

@@

First, using the full dataset, we compute the Bayesian evidence integral for each of these model choices using a nested sampling algorithm implemented in the \textsf{jaxns} package \citep{jaxns}.
The Bayesian evidence integral is defined as:
\begin{equation}
  Z(H) \equiv p(\ydata\,|\,H) = \int p(\meanparams,\hyperparams\,|\,H)\,p(\ydata\,|\,\meanparams,\hyperparams,H)\,\mathrm{d}\meanparams\,\mathrm{d}\hyperparams
\end{equation}
where $H$ indicates the modeling choices, in this case the choice of kernel function.
For better or worse\footnote{A discussion of the relative merits of different model selection algorithms is beyond the scope of this review. \suz{Sure but can we link to a more in-depth discussion of the topic for interested readers?}}, the Bayesian evidence is frequently used in the astrophysics literature as an ingredient in model selection procedures.
Nested sampling is an algorithm for numerically estimating this integral, and we apply it to produce estimates of $Z$ for each choice of kernel $H$, and plot these results in the middle panel of \autoref{fig:assessment}.
In this figure, it is clear that the squared exponential and rational quadratic kernels are indistinguishable under this metric, while the Mat\'ern-3/2 kernel is somewhat disfavoured.

@@

So far, we have primarily discussed the benefits of GP models for astrophysics, but in this section we aim to highlight some challenges faced when using GPs in practice.
The most significant challenge that has so far limited broader adoption GPs for astrophysics is their computational cost and scaling.
In the following section, we discuss some of the techniques that have been developed to scale GPs for time domain astronomy.
We also briefly discuss some contexts where GP models may overfit, or be otherwise misspecified for specific problems.

@@

The most widely-used algorithms for scalable exact GPR in time domain astrophysics are based---often not explicitly---on stochastic differential equations \citep[SDEs;][]{Sarkka:2019}, and typically require that the data be ``sortable''.
As an early example in the astrophysics literature, \citet{pr95} presented an algorithm for GPR with linear scaling with the size of the dataset, when using an exponential kernel with unevenly sampled time series data.
This method was later generalized to a wider range of covariance functions \citep{ambikasaran2015generalized, celerite, 2020AJ....160..240G, 2020A&A...638A..95D, 2022A&A...659A.182D}, and the \project{celerite}\footnote{\url{https://github.com/dfm/celerite}} and \project{S+LEAF}\footnote{\url{https://obswww.unige.ch/~delisle/spleaf/doc}} implementations have been widely-used for time domain astronomy. \dan{Add some citations to show the range.}
A qualitatively different, and independently developed approach to implementing these models uses a state-space representation for the underlying SDE to enable scalable inference \citep{2014ApJ...788...33K,  2021RNAAS...5..107J, 2022arXiv220709327M} \dan{Maybe add some citations of the use of these methods}.

@@

In our experience, a common concern that astrophysicists have when working with GPs is that these flexible models may overfit the observations, and decrease the significance of their results, especially when a GP is used as an effective model for noise sources.
While this is not an unjustified concern, we find that the risk is generally overestimated.
As discussed in \autoref{sec:gp-lsq-link}, the determinant in \autoref{eq:gp_like} penalises GP models with ``unspecific'' hyperparameters, instead preferring to explain as much of the data as possible with the mean model.
In other (but equally informal) terms, this means that a GP's modeling capacity will only be used to describe residuals that are not captured within the mean model's support.

All this being said, there do exist cases where the use of

\subsection{Model misspecification}
